A general Nullstellensatz for generalized spaces

### 2019-04-09 paper-qcoh.tex
typo
More work on the slides

\begin{defn}\label{defn:tu}
Let~$\TT$ be a geometric theory. The theory~$\ul{\TT}/U_\TT$ is the
geometric theory internal to~$\Set[\TT]$ which arises from the pulled-back
theory~$\ul{\TT}$ by adding additional constant symbols~$e_x$ of the
appropriate sorts, one for each element~$x : U_\TT$, axioms~$(\top \vdash
f(e_{x_1},\ldots,e_{x_n}) = e_{f(x_1,\ldots,x_n)})$ for each function
symbol~$f$ and~$n$-tuple of elements of~$U_\TT$ (of the appropriate sorts), and
axioms~$(\top \vdash R(e_{x_1},\ldots,e_{x_n}))$ for each relation symbol~$R$
and~$n$-tuple~$(x_1,\ldots,x_n)$ (of the appropriate sorts) such
that~$R(x_1,\ldots,x_n)$ holds for~$U_\TT$.\end{defn}

@@

As stated in Lemma~\ref{lemma:provability-in-coherent-case}, for coherent
sequents modulo a coherent theory there is no difference between provability in
geometric logic, provability$^\star$ in geometric logic and provability in
coherent logic. However, while geometric logic is powerful enough to express
provability$^\star$ in geometric logic, coherent logic is not powerful enough
to express provability in coherent logic. ``Coherent logic cannot eat itself.''
Hence the the analogue of Theorem~\ref{thm:characterization} cannot be
formulated for coherent logic.

>>>>

\begin{defn}\label{defn:tu}
Let~$\TT$ be a geometric theory. The theory~$\ul{\TT}/U_\TT$ is the
geometric theory internal to~$\Set[\TT]$ which arises from the pulled-back
theory~$\ul{\TT}$ by adding additional constant symbols~$e_x$ of the
appropriate sorts, one for each element~$x : U_\TT$, axioms~$(\top \vdash
f(e_{x_1},\ldots,e_{x_n}) = e_{f(x_1,\ldots,x_n)})$ for each function
symbol~$f$ and~$n$-tuple of elements of~$U_\TT$ (of the appropriate sorts), and
axioms~$(\top \vdash R(e_{x_1},\ldots,e_{x_n}))$ for each relation symbol~$R$
and~$n$-tuple~$(x_1,\ldots,x_n)$ (of the appropriate sorts) such
that~$R(x_1,\ldots,x_n)$.\end{defn}

@@

As stated in Lemma~\ref{lemma:provability-in-coherent-case}, for coherent
sequents modulo a coherent theory there is no difference between provability in
geometric logic, provability$^\star$ in geometric logic and provability in
coherent logic. However, while geometric logic is powerful enough to express
provability$^\star$ in geometric logic, coherent logic is not powerful enough
to express provability in coherent logic. ``Coherent logic cannot eat itself.''
Hence the analogue of Theorem~\ref{thm:characterization} cannot be
formulated for coherent logic.

### 2019-04-09 slides-birmingham2019.tex
More work on the slides

\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{ifthen}
\usepackage[normalem]{ulem}
\usepackage{soul}\setul{0.3ex}{}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing}
\hypersetup{colorlinks=true}

@@

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\renewcommand{\C}{\mathcal{C}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\CC}{\mathbb{C}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\Ring}{\mathrm{Ring}}
\newcommand{\LocRing}{\mathrm{LocRing}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ull}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
\newcommand{\Ll}{\vcentcolon\!\Longleftrightarrow}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}
\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates

@@

  The \hil{generic local ring} validates:
  \begin{enumerate}
    \item $\neg \forall x\?U_\TT\_ \neg\neg(x = 0)$ (as~$1 \neq 0$).
    \item $\forall a_0,\ldots,a_{n-1}\?U_\TT\_ \neg\neg \exists x\?U_\TT\_
      x^n + a_{n-1}x^{n-1} + \cdots + a_0 x^0 = 0$.
    \item \justifying Let~$\Delta = \{ \varepsilon \? U_\TT \,|\, \varepsilon^2 = 0 \}$.
    For any map~$f : \Delta \to U_\TT$, there are unique elements~$a,b \?
    U_\TT$ such that~$f(\varepsilon) = a + b\varepsilon$ for all~$\varepsilon
    \? \Delta$.
%   \item For any finitely presented~$U_\TT$-algebra~$A$,
%   the canonical map~$A \to (U_\TT)^{\Hom_{U_\TT}(A,U_\TT)}$
%   is an isomorphism.
  \end{enumerate}
\end{frame}

@@

\begin{frame}{Internal conservativity}
  The generic model~$U_\TT$ is conservative in the sense that:
  \begin{multline*}
    \qquad\textit{For any geometric sequent~$\sigma$,} \\
    \textit{if $\Set[\TT] \models \speak{$U_\TT \models \sigma$}$,
    then~$\TT$ proves~$\sigma$}.\qquad
  \end{multline*}
  But~$U_\TT$ \hil{does not believe} that it is conservative in the sense that
  \hil{neither}
  \begin{multline*}
    \qquad\textit{For any geometric sequent~$\sigma$,} \\
    \Set[\TT] \models (\speak{$U_\TT \models \sigma$} \Rightarrow
    \speak{$\TT$ proves~$\sigma$}) \qquad
  \end{multline*}
  \hil{nor}
  \begin{multline*}
    \qquad\Set[\TT] \models \ulcorner\text{For any geometric sequent~$\sigma$,} \\
      \text{if~$U_\TT \models \sigma$ then~$\TT$
      proves~$\sigma$}\urcorner\qquad
  \end{multline*}
  hold.
\end{frame}

@@

  \justifying
  \textbf{The topos-theoretic Nullstellensatz.}
  Internally to~$\Set[\TT]$, for any geometric$^\star$ sequent~$\sigma$ over
  the signature of~$\underline{\TT}/U_\TT$,
  \[ \text{if~$U_\TT \models \sigma$, then~$\underline{\TT}/U_\TT$
    proves$^\star$~$\sigma$.}
  \]
\end{frame}

>>>>

\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{stmaryrd}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{ifthen}
\usepackage[normalem]{ulem}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes,shapes.callouts,shapes.arrows,patterns,fit,backgrounds,decorations.pathmorphing}
\hypersetup{colorlinks=true}

@@

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\renewcommand{\C}{\mathcal{C}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\CC}{\mathbb{C}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\Ring}{\mathrm{Ring}}
\newcommand{\LocRing}{\mathrm{LocRing}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
\newcommand{\Ll}{\vcentcolon\!\Longleftrightarrow}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}
\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates

@@

  The \hil{generic local ring} validates:
  \begin{enumerate}
    \item $\neg \forall x\?U_\TT\_ \neg\neg(x = 0)$.
    \item $\forall a_0,\ldots,a_{n-1}\?U_\TT\_ \neg\neg \exists x\?U_\TT\_
      x^n + a_{n-1}x^{n-1} + \cdots + a_0 x^0 = 0$.
    \item \justifying Let~$\Delta = \{ \varepsilon \? U_\TT \,|\, \varepsilon^2 = 0 \}$.
    For any map~$f : \Delta \to U_\TT$, there are unique elements~$a,b \?
    U_\TT$ such that~$f(\varepsilon) = a + b\varepsilon$ for all~$\varepsilon
    \? \Delta$.
%   \item For any finitely presented~$U_\TT$-algebra~$A$,
%   the canonical map~$A \to (U_\TT)^{\Hom_{U_\TT}(A,U_\TT)}$
%   is an isomorphism.
  \end{enumerate}
\end{frame}

@@

\begin{frame}{Internal conservativity}
  The generic model~$U_\TT$ is conservative in the sense that
  \begin{multline*}
    \quad\textit{for any geometric sequent~$\sigma$,} \\
    \textit{if $\Set[\TT] \models \speak{$\sigma$ holds for~$U_\TT$}$,
    then~$\TT$ proves~$\sigma$}.\quad
  \end{multline*}
  But~$U_\TT$ \hil{does not believe} that it's conservative in the sense that
  \begin{multline*}
    \quad\Set[\TT] \models \ulcorner\text{for any geometric sequent~$\sigma$,} \\
      \text{if~$\sigma$ holds for~$U_\TT$, then~$\ul{\TT}$
      proves~$\sigma$}\urcorner\quad
  \end{multline*}
  does \hil{not} hold.

  \textbf{Example.} Let~$\TT$ be the theory of rings. Then
  \begin{align*}
    \Set[\TT] &\models \neg(\speak{$\ul{\TT}$ proves $(\top \vdash 1 + 1 =
    0)$}) \\
    \text{but}\quad \Set[\TT] & \not\models \neg(1 + 1 = 0).
  \end{align*}
\end{frame}

@@

%  \textbf{Definition.} The theory~$\ul{\TT}/U_\TT$ is the
%  internal geometric theory which arises from~$\ul{\TT}$ by adding:
%  \begin{enumerate}
%    \justifying
%    \item constant symbols~$e_x$, one for each element~$x : U_\TT$,
%    \item axioms~$(\top \vdash f(e_{x_1},\ldots,e_{x_n}) = e_{f(x_1,\ldots,x_n)})$ for each function
%    symbol~$f$ and~$n$-tuple~$(x_1,\ldots,x_n) \in (U_\TT)^n$,
%    \item axioms~$(\top \vdash R(e_{x_1},\ldots,e_{x_n}))$ for each relation symbol~$R$
%    and~$n$-tuple~$(x_1,\ldots,x_n) \in (U_\TT)^n$ such
%    that~$R(x_1,\ldots,x_n)$.
%  \end{enumerate}

  \textbf{The false naive topos-theoretic Nullstellensatz.}
  Internally to~$\Set[\TT]$, for any geometric sequent~$\sigma$ over
  the signature of~$\underline{\TT}$,
  \[ \text{if~$\sigma$ holds for~$U_\TT$, then~$\underline{\TT}$
    proves~$\sigma$.}
  \]

  \textbf{The topos-theoretic Nullstellensatz.}
  Internally to~$\Set[\TT]$, for any geometric$^\star$ sequent~$\sigma$ over
  the signature of~$\underline{\TT}/U_\TT$,
  \[ \text{if~$\sigma$ holds for~$U_\TT$, then~$\underline{\TT}/U_\TT$
    proves$^\star$~$\sigma$.}
  \]
\end{frame}
